Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-26 paper/ms.tex
an intro to gps

% Projects:
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\kepler}{\project{Kepler}}

@@

\section{Gaussian processes in one-dimension}

\citep{Rasmussen:2006}

>>>>

% Projects:
\newcommand{\project}[1]{\textsf{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\genrp}{\project{genrp}}

@@

\section{Gaussian processes}

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models parameterized by a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{ij} = k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
CITE}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
\dfmtodo{CITE}) algorithm to sample from the posterior probability density.

The application of GP models is generally limited to small datasets because
the computational cost of computing the inverse and determinant of the matrix
$K_\bvec{\alpha}$ scales as the cube of the number of data points $N$,
$\mathcal{O}(N^3)$.
This means that for large datasets, every evaluation of the likelihood will
quickly become computationally intractable.
In this case, standard non-linear optimization or MCMC will no longer be
practical inference methods.

In the following Section, we present a method of improving this scaling that
we call the \genrp\ method.
The \genrp\ method requires using a specific model for the covariance
$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ and it has several limitations.
The method can only be applied to one-dimensional datasets.
When we say ``one-dimensional'' here, it means that the \emph{input
coordinates} $\bvec{x}_i$ are scalar, $\bvec{x}_i \equiv t_i$.\footnote{We are
using $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the \genrp\
method can be applied to \emph{any} one-dimensional dataset.}
Furthermore, the covariance function for the \genrp\ method is ``stationary''.
This means that the function $k_\bvec{\alpha}(t_i,\,t_j)$ is only a function
of $\tau_{ij} \equiv |t_i - t_j|$.

\section{The genrp model}
